\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-02-19 Writeup/kohler_clarke_2021.tex
updated abstract

\begin{abstract}
Symmetries are present at many scales in images of natural scenes. A large body of literature has demonstrated contributions of symmetry to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in exactly 17 distinct ways. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The goal of the current study is to provide a complete description of responses to symmetry in the human visual system, by collecting both brain and behavioral data using all 17 groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that this hierarchy is preserved almost perfectly in both behavior and brain activity: A multi-level Bayesian GLM indicates that for most of the 63 subgroup relationships, subgroups produce lower amplitude responses in visual cortex (posterior probability: $>0.95$ for 56 of 63) and require longer presentation durations to be reliably detected (posterior probability: $>0.95$ for 49 of 63). This systematic pattern is seen only in visual cortex and only in components of the brain response known to be symmetric-specific. Our results show that representations of symmetries in the human brain are precise and rich in detail, and that this precision is reflected in behavior. These findings expand our understanding of symmetry perception, and open up new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

Symmetries are present at many scales in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

>>>>

\begin{abstract}
Symmetries are present at many scales in images of natural scenes. A large body of literature has demonstrated contributions of symmetry to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in exactly 17 distinct ways. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images and have recently found use in the vision science community as an ideal stimulus set for studying the perception of symmetries in textures. The goal of the current study is to provide a more comprehensive description of responses to symmetry in the human visual system, by collecting both brain imaging (Steady-State Visual Evoked Potentials measured using high-density EEG) and behavioral (symmetry detection thresholds) data using the entire set of wallpaper groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that this hierarchy is preserved almost perfectly in both behavior and brain activity: A multi-level Bayesian GLM indicates that for most of the 63 subgroup relationships, subgroups produce lower amplitude responses in visual cortex (posterior probability: $>0.95$ for 56 of 63) and require longer presentation durations to be reliably detected (posterior probability: $>0.95$ for 49 of 63). This systematic pattern is seen only in visual cortex and only in components of the brain response known to be symmetric-specific. Our results show that representations of symmetries in the human brain are precise and rich in detail, and that this precision is reflected in behavior. These findings expand our understanding of symmetry perception, and open up new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

Symmetries are abundant in natural and man-made environments, due to a complex interplay of physical forces that govern pattern formation in nature. Symmetrical patterns have been created and appreciated by human cultures throughout history and since the gestalt movement of the early 20th century, symmetry has been recognized as important for visual perception. Symmetry contributes to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four symmetries can be combined in 17 distinct ways, \textit{the wallpaper groups} \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four of the wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect was also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.
